---
title: "Rural Definitions Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rural Definitions Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

```{r libraries}
library(ruraldefinitions)
library(cori.charts)
library(tidyverse)
library(sf)
library(tigris)
library(here)

load_fonts()
options(tigris_use_cache = TRUE, tigris_progress_bar = FALSE)
```

## Parameters

Modify these blocks to change colors and output paths.

```{r params-output}
export_dir  <- here::here("export")
chart_w     <- 10
chart_h     <- 6
```

```{r params-colors}
# Brand anchor colors — all palettes are derived from these two
color_rural    <- cori_colors[["Emerald"]]       # "#00825B"
color_nonrural <- cori_colors[["Dark Purple"]]   # "#211448"

# CBSA 2023 (3 classes) -------------------------------------------------------
colors_cbsa <- c(
  "Metro"    = color_nonrural,
  "Micro"    = "#086656",
  "Non-CBSA" = color_rural
)

# RUCC 2023 (9 codes, 1 = most urban → 9 = most rural) -----------------------
colors_rucc <- setNames(
  colorRampPalette(c(color_nonrural, color_rural))(9),
  as.character(1:9)
)

# NCHS 2023 (6 codes, 1 = most urban → 6 = most rural) -----------------------
colors_nchs <- setNames(
  colorRampPalette(c(color_nonrural, color_rural))(6),
  as.character(1:6)
)

# UIC 2024 (9 codes, 1 = most urban → 9 = most rural) ------------------------
colors_uic <- setNames(
  colorRampPalette(c(color_nonrural, color_rural))(9),
  as.character(1:9)
)

# RUCA 2020 (codes 1–10, 99 = water-only tracts) -----------------------------
colors_ruca <- c(
  setNames(colorRampPalette(c(color_nonrural, color_rural))(10), as.character(1:10)),
  "99" = "#FFFFFF"
)

# CORI 2020 (3 classes) -------------------------------------------------------
colors_cori <- c(
  "Metro"                = color_nonrural,
  "Nonmetro/RUCA 4+ carveouts"             = color_rural
)
```

## Load Geometries

County and state boundaries are loaded once and reused across all county-level
maps. Tract boundaries are loaded separately for the tract-level maps.

```{r geom-county}
counties_sf <- tigris::counties(cb = TRUE, year = 2023) %>%
  tigris::shift_geometry() %>%
  filter(GEOID <= "57") %>%
  sf::st_transform(crs = 5070)

states_sf <- tigris::states(cb = TRUE, year = 2023) %>%
  tigris::shift_geometry() %>%
  filter(GEOID <= "57") %>%
  sf::st_transform(crs = 5070)

bbox <- sf::st_bbox(states_sf)
```

```{r geom-tract}
state_fips <- tigris::fips_codes %>%
  distinct(state_code) %>%
  filter(state_code <= "56") %>%
  pull(state_code)

tracts_sf <- map_dfr(
  state_fips,
  ~tigris::tracts(state = .x, cb = TRUE, year = 2020, progress_bar = FALSE)
) %>%
  tigris::shift_geometry() %>%
  sf::st_transform(crs = 5070)
```

## Maps

### CBSA 2023

Core-Based Statistical Areas (OMB). Three classes: Metro, Micro, and Non-CBSA
(nonmetro).

```{r map-cbsa}
map_cbsa <- counties_sf %>%
  left_join(cbsa_2023, by = c("GEOID" = "geoid")) %>%
  mutate(rural_def = factor(rural_def, levels = names(colors_cbsa))) %>%
  group_by(rural_def) %>%
  summarise(geometry = sf::st_union(geometry), .groups = "drop")

fig_cbsa <- ggplot() +
  geom_sf(data = map_cbsa, aes(fill = rural_def), color = NA, linewidth = 0) +
  geom_sf(data = states_sf, fill = NA, color = "black", linewidth = 0.3) +
  scale_fill_manual(name = NULL, values = colors_cbsa, na.value = "#ffffff") +
  coord_sf(
    crs = sf::st_crs(5070),
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  labs(
    title = "Core-Based Statistical Areas (CBSA), 2023",
    subtitle = "County-level OMB metropolitan/micropolitan delineation",
    caption = "Source: U.S. Office of Management and Budget, 2023 delineations."
  ) +
  theme_cori_map() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.key.width = unit(20, "pt"),
    legend.key.height = unit(20, "pt"),
    legend.spacing.x = unit(8, "pt"),
    legend.text = element_text(size = 10),
    plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.2)
  )

dir.create(export_dir, showWarnings = FALSE, recursive = TRUE)

save_plot(fig_cbsa, file.path(export_dir, "map_cbsa_2023.png"),
          chart_height = chart_h, chart_width = chart_w)
save_plot(fig_cbsa, file.path(export_dir, "map_cbsa_2023.svg"),
          chart_height = chart_h, chart_width = chart_w, add_logo = FALSE)

fig_cbsa
```

### RUCC 2023

USDA Rural-Urban Continuum Codes. Nine codes distinguishing metro counties by
population size and nonmetro counties by degree of urbanization and adjacency to
a metro area.

```{r map-rucc}
# Short labels keyed by numeric code
rucc_labels <- c(
  "1" = "Metro - 1M",
  "2" = "Metro 250K - 1M",
  "3" = "Metro <250K",
  "4" = "Nonmetro - 20K, adj",
  "5" = "Nonmetro - 20K, not adj",
  "6" = "Nonmetro 5 - 20K, adj",
  "7" = "Nonmetro 5 - 20K, not adj",
  "8" = "Nonmetro <5K, adj",
  "9" = "Nonmetro <5K, not adj"
)

map_rucc <- counties_sf %>%
  left_join(rucc_2023, by = c("GEOID" = "geoid")) %>%
  mutate(
    code  = str_extract(rural_def, "^\\d+"),
    label = factor(rucc_labels[code], levels = rucc_labels)
  ) %>%
  group_by(label) %>%
  summarise(geometry = sf::st_union(geometry), .groups = "drop")

colors_rucc_labeled <- setNames(colors_rucc, rucc_labels)

fig_rucc <- ggplot() +
  geom_sf(data = map_rucc, aes(fill = label), color = NA, linewidth = 0) +
  geom_sf(data = states_sf, fill = NA, color = "black", linewidth = 0.3) +
  scale_fill_manual(name = NULL, values = colors_rucc_labeled, na.value = "#ffffff") +
  coord_sf(
    crs = sf::st_crs(5070),
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  labs(
    title = "Rural-Urban Continuum Codes (RUCC), 2023",
    subtitle = "USDA ERS county-level classification by metro size and nonmetro urbanization",
    caption = "Source: USDA Economic Research Service, 2023 Rural-Urban Continuum Codes."
  ) +
  theme_cori_map() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.key.width = unit(20, "pt"),
    legend.key.height = unit(20, "pt"),
    legend.spacing.x = unit(6, "pt"),
    legend.text = element_text(size = 8),
    plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.2)
  )

save_plot(fig_rucc, file.path(export_dir, "map_rucc_2023.png"),
          chart_height = chart_h, chart_width = chart_w)
save_plot(fig_rucc, file.path(export_dir, "map_rucc_2023.svg"),
          chart_height = chart_h, chart_width = chart_w, add_logo = FALSE)

fig_rucc
```

### NCHS 2023

National Center for Health Statistics Urban-Rural Classification Scheme. Six
levels subdividing OMB metro areas by size and distinguishing two nonmetro
classes (micropolitan and noncore).

```{r map-nchs}
nchs_labels <- c(
  "1" = "Large central metro",
  "2" = "Large fringe metro",
  "3" = "Medium metro",
  "4" = "Small metro",
  "5" = "Micropolitan",
  "6" = "Noncore"
)

map_nchs <- counties_sf %>%
  left_join(nchs_2023, by = c("GEOID" = "geoid")) %>%
  mutate(
    code  = str_extract(rural_def, "^\\d+"),
    label = factor(nchs_labels[code], levels = nchs_labels)
  ) %>%
  group_by(label) %>%
  summarise(geometry = sf::st_union(geometry), .groups = "drop")

colors_nchs_labeled <- setNames(colors_nchs, nchs_labels)

fig_nchs <- ggplot() +
  geom_sf(data = map_nchs, aes(fill = label), color = NA, linewidth = 0) +
  geom_sf(data = states_sf, fill = NA, color = "black", linewidth = 0.3) +
  scale_fill_manual(name = NULL, values = colors_nchs_labeled, na.value = "#ffffff") +
  coord_sf(
    crs = sf::st_crs(5070),
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  labs(
    title = "NCHS Urban-Rural Classification Scheme, 2023",
    subtitle = "Six-level county classification from the National Center for Health Statistics",
    caption = "Source: National Center for Health Statistics, 2023 Urban-Rural Classification Scheme."
  ) +
  theme_cori_map() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.key.width = unit(20, "pt"),
    legend.key.height = unit(20, "pt"),
    legend.spacing.x = unit(8, "pt"),
    legend.text = element_text(size = 10),
    plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.2)
  )

save_plot(fig_nchs, file.path(export_dir, "map_nchs_2023.png"),
          chart_height = chart_h, chart_width = chart_w)
save_plot(fig_nchs, file.path(export_dir, "map_nchs_2023.svg"),
          chart_height = chart_h, chart_width = chart_w, add_logo = FALSE)

fig_nchs
```

### UIC 2024

USDA Urban Influence Codes. Nine codes classifying counties by population,
degree of urbanization, and adjacency to metro areas.

```{r map-uic}
uic_labels <- c(
  "1" = "Large metro",
  "2" = "Micro, adj large metro",
  "3" = "Noncore, adj large metro",
  "4" = "Small metro",
  "5" = "Micro, adj small metro",
  "6" = "Noncore, adj small metro",
  "7" = "Micro, not adj to metro",
  "8" = "Noncore, not adj (>5K town)",
  "9" = "Noncore, not adj (<5K no town)"
)

map_uic <- counties_sf %>%
  left_join(uic_2024, by = c("GEOID" = "geoid")) %>%
  mutate(
    code  = str_extract(rural_def, "^\\d+"),
    label = factor(uic_labels[code], levels = uic_labels)
  ) %>%
  group_by(label) %>%
  summarise(geometry = sf::st_union(geometry), .groups = "drop")

colors_uic_labeled <- setNames(colors_uic, uic_labels)

fig_uic <- ggplot() +
  geom_sf(data = map_uic, aes(fill = label), color = NA, linewidth = 0) +
  geom_sf(data = states_sf, fill = NA, color = "black", linewidth = 0.3) +
  scale_fill_manual(name = NULL, values = colors_uic_labeled, na.value = "#ffffff") +
  coord_sf(
    crs = sf::st_crs(5070),
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  labs(
    title = "Urban Influence Codes (UIC), 2024",
    subtitle = "USDA ERS county classification by urbanization and metro adjacency",
    caption = "Source: USDA Economic Research Service, 2024 Urban Influence Codes."
  ) +
  theme_cori_map() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.key.width = unit(20, "pt"),
    legend.key.height = unit(20, "pt"),
    legend.spacing.x = unit(6, "pt"),
    legend.text = element_text(size = 8),
    plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.2)
  )

save_plot(fig_uic, file.path(export_dir, "map_uic_2024.png"),
          chart_height = chart_h, chart_width = chart_w)
save_plot(fig_uic, file.path(export_dir, "map_uic_2024.svg"),
          chart_height = chart_h, chart_width = chart_w, add_logo = FALSE)

fig_uic
```

### RUCA 2020

USDA Rural-Urban Commuting Area Codes. Ten tract-level codes based on
urbanization, population density, and commuting patterns. Code 99 indicates
water-only tracts.

```{r map-ruca}
ruca_labels <- c(
  "1"  = "Metro core",
  "2"  = "Metro high commuting",
  "3"  = "Metro low commuting",
  "4"  = "Micro core",
  "5"  = "Micro high commuting",
  "6"  = "Micro low commuting",
  "7"  = "Small town core",
  "8"  = "Small town high commuting",
  "9"  = "Small town low commuting",
  "10" = "Rural",
  "99" = "Water only"
)

ruca_raw <- get_definition("ruca", 2020)

map_ruca <- tracts_sf %>%
  left_join(ruca_raw, by = c("GEOID" = "geoid")) %>%
  mutate(
    code  = str_extract(rural_def, "^\\d+"),
    label = factor(ruca_labels[code], levels = ruca_labels)
  ) %>%
  filter(code != "99", !is.na(code)) %>%
  group_by(label) %>%
  summarise(geometry = sf::st_union(geometry), .groups = "drop")

colors_ruca_labeled <- setNames(colors_ruca, ruca_labels)

fig_ruca <- ggplot() +
  geom_sf(data = map_ruca, aes(fill = label), color = NA, linewidth = 0) +
  geom_sf(data = states_sf, fill = NA, color = "black", linewidth = 0.3) +
  scale_fill_manual(name = NULL, values = colors_ruca_labeled, na.value = "#ffffff") +
  coord_sf(
    crs = sf::st_crs(5070),
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  labs(
    title = "Rural-Urban Commuting Area Codes (RUCA), 2020",
    subtitle = "Tract-level classification based on urbanization, density, and commuting flows",
    caption = "Source: USDA Economic Research Service, 2020 RUCA Codes."
  ) +
  theme_cori_map() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.key.width = unit(20, "pt"),
    legend.key.height = unit(20, "pt"),
    legend.spacing.x = unit(6, "pt"),
    legend.text = element_text(size = 8),
    plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.2)
  )

save_plot(fig_ruca, file.path(export_dir, "map_ruca_2020.png"),
          chart_height = chart_h, chart_width = chart_w)
save_plot(fig_ruca, file.path(export_dir, "map_ruca_2020.svg"),
          chart_height = chart_h, chart_width = chart_w, add_logo = FALSE)

fig_ruca
```

### CORI 2020

CORI's hybrid definition combining nonmetropolitan counties (CBSA 2023) with
rural tracts in metropolitan counties (RUCA 2020 codes 4+).

```{r map-cori}
cori_raw <- get_definition("cori", 2020)

map_cori <- tracts_sf %>%
  left_join(cori_raw, by = c("GEOID" = "geoid")) %>%
  filter(!is.na(rural_def)) %>%
  mutate(
    display = case_when(
      rural_def == "Metro county"                ~ "Metro",
      rural_def %in% c("Nonmetro county",
                       "Rural tract in metro county") ~ "Nonmetro/RUCA 4+ carveouts"
    ),
    display = factor(display, levels = names(colors_cori))
  ) %>%
  group_by(display) %>%
  summarise(geometry = sf::st_union(geometry), .groups = "drop")

fig_cori <- ggplot() +
  geom_sf(data = map_cori, aes(fill = display), color = NA, linewidth = 0) +
  geom_sf(data = states_sf, fill = NA, color = "black", linewidth = 0.3) +
  scale_fill_manual(name = NULL, values = colors_cori, na.value = "#ffffff") +
  coord_sf(
    crs = sf::st_crs(5070),
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  labs(
    title = "CORI Rural Definition, 2020",
    subtitle = "Nonmetro counties (CBSA 2023) plus rural tracts (4+) in metro counties (RUCA 2020)",
    caption = "Source: Center on Rural Innovation."
  ) +
  theme_cori_map() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.key.width = unit(20, "pt"),
    legend.key.height = unit(20, "pt"),
    legend.spacing.x = unit(8, "pt"),
    legend.text = element_text(size = 10),
    plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.2)
  )

save_plot(fig_cori, file.path(export_dir, "map_cori_2020.png"),
          chart_height = chart_h, chart_width = chart_w)
save_plot(fig_cori, file.path(export_dir, "map_cori_2020.svg"),
          chart_height = chart_h, chart_width = chart_w, add_logo = FALSE)

fig_cori
```
